Replacing standard galaxy profiles with \\ mixtures of Gaussians

### 2012-10-22 documents/mixtures/mixture_models.tex
badness regularization, advice to users of various kinds

% to-do
% -----
% - URGENT: badness has another term in it -- the regularization term
% - DSTN: make figures showing use of the models in a real situation; see last figure
% - does the Lupton paper give the profiles, or should we cite the code directly?
% - make sure the technical issues about convolution and interpolation don't exist *only* in the figure caption
% - send to Jim Bosch and Kevin Bundy for comments

@@

To elaborate, a mixture of multivariate Gaussians---a linear
superposition---can be used to represent any reasonable distribution
in any number of dimensions to any reasonable precision.  Convolution
(or correlation) by any other distribution that has also been
represented by a mixture of multivariate Gaussians creates a new
mixture of Gaussians with simply adjusted amplitudes, means, and
variance tensors.  The ubiquity of convolution operations in astronomy
suggests the widespread adoption of mixture-of-Gaussian modeling.  We
have pioneered this in the area of distribution modeling in large
numbers of dimensions (\citealt{xd, xdqso, xdqsoz}), where convolution
occurs because the true or noise-free distribution is convolved with
the noise before being observed.  Here we are going to capitalize on
the convolution properties of mixtures of Gaussians in modeling galaxy
morphologies in imaging data, where the true or
high-angular-resolution intensity field is convolved with the
point-spread function (PSF) before being observed.  This is
particularly useful in data sets in which the PSF itself has also been
modeled as a mixture of Gaussians, which is not uncommon (for example,
the \project{Sloan Digital Sky Survey} imaging pipelines
make approximate mixture-of-Gaussian PSF models for
every imaging field; \citealt{lupton}).

@@

... insert here a paragraph about the \emph{wisdom} of fitting these
standard profiles: When is it scientifically okay?...

@@

We need good performance in two-dimensional image synthesis (for
fitting) because, with \thetractor\ (Lang \etal, forthcoming), we are
building a comprehensive model of all the imaging data we have; this
will have models of many millions of galaxies in imaging that contains
on the order of $10^{13}$ pixels.  We are basing our models on the
\project{Sloan Digital Sky Survey} (\sdss) Catalog galaxy models,
which only include exponential, de~Vaucouleurs, and composite (mixture
of the two) radial profiles.  In detail, in fact, the \sdss\ Catalog
models are small modifications of these (\citealt{lupton}); details
below.  So our goal here is to provide replacements for these models
in order to improve the performance of galaxy image modeling and
analysis software.  Some of the models we present have been used
previously in tools for precise photometry (Bundy \etal, forthcoming);
they could also be used to speed other image-fitting systems, like the
very successful \project{GALFIT} (\citealt{galfit}).

@@

In detail, the chi-squared objective we minimize---the
\emph{badness}---is a squared residual between the exact profile
function $Q(\spos)$ and its mixture-of-Gaussian approximation.  It is
designed to be equivalent to a chi-squared statistic in a
homoskedastic two-dimensional image of the profile taken with
extremely high angular resolution (pixels of size 0.001 the half-light
radius) and vanishing point-spread function.  Quantitatively the
badness is defined to be the mean squared residual in the $Q(\spos)$
functions, which are normalized to have unit intensity at the
half-light radius, averaged over a two-dimensional circular region in
the $\spos$ plane centered on the (circularly symmetric) profile and
extending out to radius $\xi_{\max}$.  We use $\xi_{\max}=8$ for all
profiles except the $\lux$ profile, for which we use $\xi_{\max}=4$.
In practice, the badness is computed in a one-dimensional numerical
integral but the integral is weighted in radius (weight increasing
linearly with radius) to make it equivalent to the two-dimensional
chi-squared.

@@

... make recommendations: Use the $M^{\lux}=6$ lux and the $M^{\luv}=8$
luv!...

@@

... pegagogical figure demonstrating these operations and their results...

@@

\begin{thebibliography}{70}
\bibitem[Bendinelli(1991)]{bendinelli}
Bendinelli,~O., 1991, \apj, 366, 599
\bibitem[Bendinelli \& Parmeggiani(1995)]{bendinelli2}
Bendinelli,~O. \& Parmeggiani,~G., 1995, \aj, 109, 572
\bibitem[Bovy \etal(2011a)]{xd}
Bovy,~J., Hogg,~D.~W., \& Roweis,~S., 2011, Ann. Appl. Stat., 5, 1657
\bibitem[Bovy \etal(2011b)]{xdqso}
Bovy,~J., \etal, 2011, \apj, 729, 141
\bibitem[Bovy \etal(2012)]{xdqsoz}
Bovy,~J. \etal, 2012, \aj, 749, 41
\bibitem[Cappellari(2002)]{cappellari}
Cappellari,~M., 2002, \mnras, 333, 400
\bibitem[Ciotti \& Bertin(1999)]{ciotti}
Ciotti,~L. \& Bertin,~G., 1999, \aa, 352, 447
\bibitem[de~Vaucouleurs(1948)]{dev}
de~Vaucouleurs,~G., 1948, Annales d'Astrophysique, 11, 247
\bibitem[Emsellem \etal(1994)]{emsellem}
Emsellem,~E., Monnet,~G., Bacon,~R., \& Nieto,~J.-L., 1994, \aap, 285, 739
\bibitem[Lupton \etal(2001)]{lupton}
Lupton,~R., Gunn,~J.~E., Ivezic,~Z., Knapp,~G.~R., Kent,~S.~M., \& Yasuda,~N., 2001, ASPC, 238, 269
\bibitem[Peng \etal(2002)]{galfit}
Peng,~C.~Y., Ho, L.~C., Impey,~C.~D., \& Rix,~H.-W., 2002, \aj, 124, 266
\bibitem[S{\'e}rsic(1963)]{ser}
S{\'e}rsic,~J.~L., 1963, Boletin de la Asociacion Argentina de Astronomia La Plata Argentina, 6, 41
\end{thebibliography}

>>>>

% to-do
% -----
% - HOGG: can we drop the regularization in the badness and still succeeed?  If so, let's do that (and drop it in the text).
% - DSTN: make figures showing use of the models in a real situation; see last figure
% - HOGG: does the Lupton paper give the profiles, or should we cite the code directly?
% - make sure the technical issues about convolution and interpolation don't exist *only* in the figure caption
% - HOGG: need Bundy reference
% - HOGG: make what I say about M=6 lux and M=8 lux true -- ie, edit the Tractor code!
% - send to Jim Bosch and Kevin Bundy for comments

@@

To elaborate, a mixture of multivariate Gaussians---a linear
superposition---can be used to represent any reasonable distribution
in any number of dimensions to any reasonable precision.  Convolution
(or correlation) by any other distribution that has also been
represented by a mixture of multivariate Gaussians creates a new
mixture of Gaussians with simply adjusted amplitudes, means, and
variance tensors.  The ubiquity of convolution operations in astronomy
suggests the widespread adoption of mixture-of-Gaussian modeling.  We
have pioneered this in the area of distribution modeling in large
numbers of dimensions (\citealt{xd, xdqso, xdqsoz}), where convolution
occurs because the true or noise-free distribution is convolved with
the noise before being observed.  Here we are going to capitalize on
the convolution properties of mixtures of Gaussians in modeling galaxy
morphologies in imaging data, where the true or
high-angular-resolution intensity field is convolved with the
point-spread function (PSF) before being observed.  This is always
easier with Gaussians than cuspy profiles, but is \emph{particularly}
useful in data sets in which the PSF itself has also been modeled as a
mixture of Gaussians, which is not uncommon (for example, the
\project{Sloan Digital Sky Survey} imaging pipelines make approximate
mixture-of-Gaussian PSF models for every imaging field;
\citealt{lupton}).

@@

Whenever an investigator is fitting PSF-convolved exponential or
de~Vaucouleurs profiles, the models presented here will improve code
performance.  That doesn't mean that doing such fitting is a Good
Idea.  These profiles are effective models for galaxies at low
signal-to-noise (as they are used in the \project{SDSS}) and they are
useful templates for performing consistent photometry across
morphologically diverse objects (as we use them in \citealt{bundy}).
They are sometimes used to distinguish disk and bulge light; these
uses are very prone to over-interpretation: There is no theoretical
argument and only weak observational arguments that the
rotation-supported parts of galaxies are always exponential and that
the kinematically hot components are de~Vaucouleurs.

@@

We need good performance in two-dimensional image synthesis (for
fitting) because, with \thetractor\ (Lang \etal, forthcoming), we are
building a comprehensive model of all the imaging data we have; this
will have models of many millions of galaxies in imaging that contains
on the order of $10^{13}$ pixels.  We are basing our models on the
\project{Sloan Digital Sky Survey} (\sdss) Catalog galaxy models,
which only include exponential, de~Vaucouleurs, and composite (mixture
of the two) radial profiles.  In detail, in fact, the \sdss\ Catalog
models are small modifications of these (\citealt{lupton}); details
below.  So our goal here is to provide replacements for these models
in order to improve the performance of galaxy image modeling and
analysis software.  Some of the models we present have been used
previously in tools for precise photometry (\citealt{bundy}); they
could also be used to speed other image-fitting systems, like the very
successful \project{GALFIT} (\citealt{galfit}).

@@

In detail, the chi-squared objective we minimize---the
\emph{badness}---is a squared residual between the exact profile
function $Q(\spos)$ and its mixture-of-Gaussian approximation.  It is
designed to be equivalent to a chi-squared statistic in a
homoskedastic two-dimensional image of the profile taken with
extremely high angular resolution (pixels of size 0.001 the half-light
radius) and vanishing point-spread function.  Quantitatively the
badness is defined to be the mean squared residual in the $Q(\spos)$
functions, which are normalized to have unit intensity at the
half-light radius, averaged over a two-dimensional circular region in
the $\spos$ plane centered on the (circularly symmetric) profile and
extending out to radius $\xi_{\max}$.  We use $\xi_{\max}=8$ for all
profiles except the $\lux$ profile, for which we use $\xi_{\max}=4$.
In practice, the badness is computed in a one-dimensional numerical
integral but the integral is weighted in radius (weight increasing
linearly with radius) to make it equivalent to the two-dimensional
chi-squared.  We also add to the badness a very tiny coefficient (on
the order of $10^{-3}$ of the best-fit badness) times the sum of the
variances $v_m$ for regularization.  In practice, this term doesn't
have much effect and could be dropped.

@@

In our work on \project{The Tractor}, we use the $M^{\lux}=6$ lux and
the $M^{\luv}=8$ luv profiles.  Our advice to users would be to do the
same.  We use the lux and luv over the exp and dev partly because of
their better behaviors numerically, and partly because they and we are
both part of the \project{SDSS} tradition.  These---$M^{\lux}=6$ and
$M^{\luv}=8$ are good compromises between mixture complexity ($M$) and
quality of fit.  Also, even the best-fitting late-type and early-type
galaxies deviate from exponential and de~Vaucouleurs fits by more than
do these high-quality mixture-of-Gaussian approxmations; no precision
is lost.

@@

DSTN... pegagogical figure demonstrating these operations and their results...

@@

\begin{thebibliography}{70}
\bibitem[Bendinelli(1991)]{bendinelli}
Bendinelli,~O., 1991, \apj, 366, 599
\bibitem[Bendinelli \& Parmeggiani(1995)]{bendinelli2}
Bendinelli,~O. \& Parmeggiani,~G., 1995, \aj, 109, 572
\bibitem[Bovy \etal(2011a)]{xd}
Bovy,~J., Hogg,~D.~W., \& Roweis,~S., 2011, Ann. Appl. Stat., 5, 1657
\bibitem[Bovy \etal(2011b)]{xdqso}
Bovy,~J., \etal, 2011, \apj, 729, 141
\bibitem[Bovy \etal(2012)]{xdqsoz}
Bovy,~J. \etal, 2012, \aj, 749, 41
\bibitem[Bundy \etal(2012)]{bundy}
Bundy,~K.~??
\bibitem[Cappellari(2002)]{cappellari}
Cappellari,~M., 2002, \mnras, 333, 400
\bibitem[Ciotti \& Bertin(1999)]{ciotti}
Ciotti,~L. \& Bertin,~G., 1999, \aa, 352, 447
\bibitem[de~Vaucouleurs(1948)]{dev}
de~Vaucouleurs,~G., 1948, Annales d'Astrophysique, 11, 247
\bibitem[Emsellem \etal(1994)]{emsellem}
Emsellem,~E., Monnet,~G., Bacon,~R., \& Nieto,~J.-L., 1994, \aap, 285, 739
\bibitem[Lupton \etal(2001)]{lupton}
Lupton,~R., Gunn,~J.~E., Ivezic,~Z., Knapp,~G.~R., Kent,~S.~M., \& Yasuda,~N., 2001, ASPC, 238, 269
\bibitem[Peng \etal(2002)]{galfit}
Peng,~C.~Y., Ho, L.~C., Impey,~C.~D., \& Rix,~H.-W., 2002, \aj, 124, 266
\bibitem[S{\'e}rsic(1963)]{ser}
S{\'e}rsic,~J.~L., 1963, Boletin de la Asociacion Argentina de Astronomia La Plata Argentina, 6, 41
\end{thebibliography}
